Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-07-13 text/manuscript-full-KS.qmd
copy edits

On January 9, 2019 a public research university purchased College Board Student Search Service order no. $448922$, thereby obtaining the contact information of out-of-state prospective students, who would subsequently receive brochures, postcards, emails, and targeted social media. This was a large student list purchase, yielding $122,426$ "names." At $\$0.45$ per name, the order cost $\$55,091.70$. The order specified three "search filters" to control which names were purchased: first, the 2020 high school graduating class; second, under the heading "College Board Exams," PSAT score range of 1070 - 1180; and third, under the heading "Geography", the order filtered on 67 "Geomarkets," identified by titles like "IL08 - Northwest Suburbs" and "IL09 - North Shore."[^1] The Geomarket filter carves states and metropolitan areas into smaller geographic units. Upon closer inspection, this student list purchase selected all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago," which is significantly poorer and has a higher share of Black residents. Similarly, in the DC/Maryland/Virginia metro area, the order selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets.

@@

**Correlation**. Broadly, correlation measures the extent to which two or more variables move together. Predictive algorithms are based on correlation and are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and/or to assign levels of risk to each case. @RN4975 provides the example of @RN4989, who developed a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 [p. 50] writes that, "correlation grounds big data's so-called revolutionary potential. As Wired editor Chris Anderson infamously declared \[\ldots\], big data proved that 'correlation supersedes causation, and science can advance even without coherent models, unified theories, or really any mechanistic explanation at all.'"

@@

Negative consequences of homophily
Chun (2021) discusses the negative consequences arising from homophily.Second, when we “engineer homophily” – through algorithms or policies – we make it less likely for heterophily to emerge, eliminating possibilities – of friendships, of matches – that may have occurred in the absence of such engineering.

@@

**Geomarket Filter in Student Search Service**. <!-- By 1984, College Board Geomarkets were incorporated as a search filter in the College Board Student Search Service \[CITE\]. --> Geomarkets also function as a student list search filter for the College Board's Student Search Service. A student list contains the contact information of prospective students who meet the search filter criteria (e.g., test score, GPA) specified by the university. Student lists are the fundamental input for undergraduate recruiting campaigns because purchased names -- alongside prospects who reach out on their own -- constitute the set of prospects who receive subsequent recruiting interventions (e.g., mail, email) designed to push them toward the application and enrollment stages of the “enrollment funnel.” @RN4895 reports that 86% of public colleges and 87% of private colleges purchase student lists.[^11] Historically, dominant list vendors are the College Board and ACT, which derived student list data from test-takers, but the test-optional movement, advances in technology, and surging private equity investments have contributed to new sources of student list data [@list_biz].

@@

**RQ1**. RQ1 primarily utilizes Census Data while RQ2 primarily utilizes data about student list purchases. Research question 1 is about socioeconomic and racial variation between Geomarkets in metropolitan areas and how such variation changes over time. To answer this question, we use census tract data from the U.S. Census Bureau. Utilizing a small geographic area is important for implementing a spatial merge between Census data and Geomarket shape files (described below). Census tracts are the smallest geographic area for which measures of both race/ethnicity and socioeconomic characteristics are available. Defined as statistical subdivisions of counties, census tracts boundaries are reassessed every decennial census to reflect relatively homogeneous demographic communities, typically ranging in population size from 2,500 and 8,000 residents [@Bureau2014; @Bureau1994]. Compared to zip codes, which are fixed spatial units designed primarily to demarcate route efficiency for postal services, census tracts are considered more reliable for population-based analyses.

@@

Across these data sources, we created measures of race and ethnicity, specifically the number and proportion of the census-tract population in the following categories: White, non-Hispanic; Black, non-Hispanic; Asian, non-Hispanic; Pacific Islander, non-Hispanic; American Indian and Alaskan Native, non-Hispanic; and two or more races, non-Hispanic. Measures of Asian, American Indian and Alaskan Native, two or more races, and Native Hawaiian and Pacific Islander (all non-Hispanic) were not available for the 1980 Decennial Census. We also created socioeconomic measures that approximate the predictors of market segment identified by @RN4982. These variables are median household income, percent of households below the poverty line, and people age 25+ with a BA. All measures were created at the census tract-level and at the the Geomarket-level by aggregating tract-level data to Geomarkets.

@@

[^17]: An alternative approach to dealing with spatial misalignment included creating shapefiles for Geomarkets from a 2012 R-bloggers post written by a data scientist working in the enrollment management space. Essentially, the data file was a Geomarket-zip code crosswalk that held one observation per zip code and a column that assigned each zip code to one Geomarket. Using these zip code-level data, we utilized the aggregate() function to create shapefiles for each Geomarket. Next, we utilized the st_intersection() function to assign each census tract to a Geomarket, based on the intersection of their associated spatial shapefiles. This spatial merge kept Geomarket boundaries consistent across all three Decennial Census and split census-tract shapefiles that intersected multiple Geomarkets into smaller shapefiles, each entirely contained within a single Geomarket. We replaced each variable in these shapefiles with its original value scaled by the share of the tract’s land area that lay within the Geomarket. For a particular census tract, imagine that 60% of its land area was contained in Geomarket A, 40% of its land area was contained in Geomarket B, and the census tract reports 100 people who identify as Asian, non-Hispanic. The partial spatial merge splits this census tract into two observations. The observation assigned to Geomarket A reports $.60*100 = 60$ Asian, non-Hispanic people and the observation assigned to Geomarket B reports $.40*100 = 40$ Asian, non-Hispanic people. Even under this most conservative approach to addressing spatial misalignment, overall patterns in the findings remained consistent. However, we opted to georeference the manually digitized maps directly sourced from the College Board, as the original source of the Geomarket-zip code crosswalk cited in the the R-bloggers post could not be verified.

@@

The data collection sample for this project was all public universities in California, Illinois, Minnesota, and Texas. We chose these states because they were priorities for the two foundations that funded the data collection. Utilizing public records requests to obtain quantitative data is a painstaking process, made more difficult by the onset of the COVID19 Pandemic. Initially, the majority of universities ignored or denied our request. We established a client relationship with a civil rights legal organization and leveraged this relationship to obtain pro bono representation from four law firms. Firm representation substantially increased the success of data collection. We also collected data from Arizona State University and Northern Arizona University, because we were able to obtain pro-bono legal representation for these two universities. However, we were unable to obtain legal representation for Texas because firms interested in representing us had conflicts of interest in Texas.

@@

The variables of interest are Geomarket and prospect-level characteristics, which are derived from the pretest questionnaire administered to College Board test-takers that are included in the de-identified student lists we received (e.g., HS code, HS GPA range, intended major). College Board asks students separate questions about ethnicity (Cuban, Mexican, Puerto Rican, other Hispanic, non-Hispanic, ethnicity non-response) and race (American Indian or Alaska Native, Asian, Black, Native Hawaiian or other Pacific Islander, White, race non-response), allowing students to check as many boxes as they want [@RN0010]. From these responses, a College Board "derived aggregate race/ethnicity" was re-created based on the U.S. Department of Education reporting guidelines that includes the following categories: no response; American Indian/Alaska Native; Asian; Black; Hispanic/Latino; Native Hawaiian or Other Pacific Islander; White; other; two or more races, non-Hispanic [@RN0010].[^20] Information about student socioeconomic status is limited. However, we create a measure of whether the prospect is first-generation college student based on their parents' level of education that was self-reported on the College Board questionnaire.

[^20]: Any student who selects a Hispanic ethnicity category is defined as Hispanic/Latino, regardless of the race categories they select. Additionally, non-Hispanic students who check “American Indian or Alaska Native” and another race group are defined as "two or more races, non-Hispanic."

@@

**Research design**. We utilize a multiple, quantitative case study design. Metropolitan areas are cases. Case studies deepen understanding of phenomena by comparing and contrasting across different settings [@RN4011]. Our research questions explore the socioeconomic and racial variation between Geomarkets, how such variation changes over time, and how Geomarket landscapes shape the composition of included versus excluded prospects. Our goal is not to make generalizations about Geomarket variation nationally, but to explore how different geographical models of metropolitan areas, shaped by their unique demographic characteristics, influence the composition of prospect pools. Such aims align with the purpose of case studies, which drive the exploration of research questions focused on the "how" and "why" of phenomenon in rich and nuanced contexts. We, therefore, select a small number of metropolitan areas in order to provide sufficient analytic depth while situating each case within its specific historic and contemporary sociodemographic characteristics. The [Online Appendix](https://ozanj.github.io/cb_geo/#/results) presents results for a larger number of metropolitan areas.

The central thesis of Zemsky & Oedel's (1983) Market Segment Model suggests that student demand for various colleges is a function of social class and geography, and that colleges strategically target recruiting in high socioeconomic areas. However, because Geomarkets vary significantly in their racial and class characteristics across Geomarkets, and in how colleges target these Geomarkets, a multiple case study enables us to compare how these dynamics unfold across different metropolitan areas. By treating each metropolitan area as a separate case, we explore of both within-case processes (e.g., how homophily structures recruitment in a single metropolitan area) and cross-case patterns (e.g., how the same recruitment strategy results in different racial/class outcomes across metropolitan areas).

Case selection was informed by purposeful sampling to identify "information-rich" cases that can provide knowledge about variation across Geomarkets and over time [@RN4011; @RN4549] . We considered metropolitan areas as information rich cases across three broad dimensions: 1) urban model structures, 2) geographical variation, 3) variation across racial/ethnic and socioeconomic characteristics, and 4) availability of student list orders and de-identified list data. Based on @RN4982, we identified classic urban model structures that reflected "major metropolitan area\[s\] \ldots composed of several markets, usually corresponding to the inner city, a first ring of suburbs, and an outer ring of suburbs" (p. 11-12), such as Boston, Chicago, Detroit, Baltimore, Denver, Houston, Miami, and Philadelphia. To compare and contrast with classic urban model structures, we also identified alternative urban models such as bi-modal or metroplex models that contain two centralized cores surrounded by the more classing suburban ring (e.g., Dallas-Fort Worth and San Francisco) and multi-nuclei models where urban growth develops around multiple, sometimes uncentralized and independent, metropolitan areas (e.g., Los Angeles and New York City). The third consideration addresses geographical variations in demographic populations by intentionally selecting metropolitan cases across different U.S. regions (West, Midwest, Northeast, South) to ensure resident and prospect pools mirror broader populations. Related to all previous selection dimensions, we are also interested in variation of racial and socioeconomic populations between Geomarkets within a metropolitan areas (RQ1) and how such variation shapes which prospective students are targeted by student list purchases (RQ2). We therefore considered whether or not metropolitan areas provided sufficient and varied racial and socioeconomic diversity for our research purposes.

Lastly, choice of metropolitan cases was informed by student list data availability. Although Census data are available for all metropolitan areas nationally (RQ1), we do not have good candidate student list data for all metropolitan areas (RQ2). In order to show which prospective students would have been included or excluded from the student list purchase had the purchase filtered on particular Geomarkets, the student list purchases we analyze must include all Geomarkets in a metropolitan area. This lead us to select on orders that filtered for an entire state (e.g., State filter is set to Pennsylvania), an entire metropolitan area without a Geomarket filter (CBSA filter is set to the Philadelphia metropolitan area), or an entire metropolitan area with a Geomarket filter inclusive of all Geomarkets (CBSA filter is set to the Philadelphia metropolitan area and Geomarket filter is set to all five Geomarkets in the Philadelphia area). Although we would like to analyze the D.C./Maryland/Virginia (DMV) area, the student list purchases we collected exclude "MD 5 -- Prince Georges Metropolitan" and "MD 7 -- Baltimore (Urban), which are substantially poorer and more Black than other Geomarkets in the DMV. Similarly, student list purchases targeting Cleveland exclude "OH 4 -- City of Cleveland (East)."

@@

We answer RQ2 by analyzing student list purchases that encompassed all Geomarkets in a selected metropolitan areas. Descriptive tables and graphs describe the racial and socioeconomic characteristics of prospects that would have been included/excluded had the purchase filtered on particular Geomarkets. We also create interactive maps of prospect pools by Geomarket contributions to provide a more granular visualization of analyses.

@@

@fig-chicago-maps presents the seven Geomarkets in the Chicago metropolitan area by total population across the three census periods. The maps in the figure show how Chicago Geomarkets reflect the classic urban geographical structure of an inner-city core surrounded by a ring of suburbs as described by @RN4982 [p. 12]. Based on the color gradient scale representing total population density, @fig-chicago-maps indicates northern suburb Geomarkets in the metropolitan area - Chain of Lakes, Northwest Suburbs, Northshore, Evanston and Skokie - experienced relatively little population growth from 1980 through 2020. In contrast, the most significant population increases occurred in the Western Suburbs (1.0 Million to 1.4 Million) and the South and Southwest Suburbs (less than 1 Million to 1.2 Million) Geomarkets. The City of Chicago Geomarket, forming the metropolitan area's central core, was the only locality to experience a population decline, from 3.5 Million in 1980 to 3.2 Million in 2020, despite remaining the most populous Geomarket.

These modest population shifts belie substantial change in the racial/ethnic composition within and across Geomarkets over time. @fig-chicago-rq1-race illustrates the racial and ethnic composition of each Geomarket across the three census periods. In 1980, nearly all Geomarkets were predominantly White, with proportions ranging from 84% in the South and Southwest Suburbs to approximately 96% in North Shore. In contrast, the City of Chicago had a substantially more diverse population, with White residents comprising only 52% of its total population. One data caveat is that the following race/ethnicity categories were not available in 1980: Asian, non-Hispanic; Two+ races, non-Hispanic; NHPI, non-Hispanic; AIAN non-Hispanic. By 2020, several Geomarkets that were previously predominantly White shifted to a more heterogeneous racial/ethnic distribution, including Chain of Lakes, Northwest Suburbs, Evanston and Skokie, Western Suburbs, and South and Southwest Suburbs. For many Geomarkets, this balance in racial/ethnic composition was the result of large increases in the Latinx population from 1980 to 2020, which ranged from a 7 percentage point increase for the Evanston and Skokie Geomarket to a 20 percentage point increase for Chain of Lakes. Some Geomarkets also experienced a substantial increase in Black populations (South and Southwest Suburbs) and in Asian populations (Northwest Suburbs and Evanston and Skokie). @fig-chicago-rq1-race also indicates that, although the North Shore Geomarket underwent some changes in racial and ethnic composition, its overall demographic patterns remained relatively the same across the three census periods. City of Chicago experienced relatively modest declines in its Black population (35% in 1980 to 26% in 2020), alongside increases in its Hispanic (13% in 1980 to 30% in 2020) and Asian Pacific Islander populations (4% in 2000 to 7% in 2020), further substantiating the Geomarket as the most racially/ethnic diverse in the metropolitan area. <!-- For instance, North Shore experienced some increases in Hispanic (2% in 1980 to 8% in 2020), Asian Pacific Islander (7% in 2000 to 10% in 2020), and Black (2% in 1980 to 3% in 2020) populations yet remained a predominantly White Geomarket (74% in 2020). City of Chicago also experienced relatively modest declines in its Black population (37% in 1980 to 33% in 2020), alongside increases in its Hispanic (15% in 1980 to 26% in 2020) and Asian Pacific Islander populations (4% in 2000 to 6% in 2020), further substantiating the Geomarket as the most racially/ethnic diverse in the metropolitan area. -->

The interactive maps in the [Online Appendix](https://ozanj.github.io/cb_geo/#/research-question-1-15) can show these shifts at the Census Tract-level. For example, in 1980 Black people were concentrated in southern tracks of the City of Chicago Geomarket. As described by scholarship on gentrification in Chicago [@ewing2024_nofuture; @snidal2022_exodus]. However, in 2000 and 2020 Black populations shift towards northeast tracks of the South and Southwest Suburbs Geomarket

We explore socioeconomic demographics -- median income, percent living below poverty, and educational attainment -- within and across Geomarkets over time in @fig-chicago-rq1-ses. The primary takeaways are the City of Chicago -- and to a somewhat lesser degree th South and Southwest Suburbs - stands out as having lower median income, higher poverty rates, and lower BA attainment than other Geomarkets. By contrast, the North Shore stands out as being particularly affluent and highly educated. Overall, median income both within and across Geomarkets in the Chicago metropolitan area remained relatively stable across the three census periods. For instance, in 1980, median household income across the Chicago metropolitan area ranged from a high of \$165,000 in the North Shore Geomarket to a low of \$68,000 in the City of Chicago Geomarket. However, median incomes in all other Geomarkets were closer to the upper thresholds of affluence, such as \$125,000 in Evanston and Skokie, \$117,000 in the Northwest Suburbs, \$108,000 in the Western Suburbs, \$104,000 in Chain of Lakes, and \$99,000 in the South and Southwest Suburbs. While overall income patterns across Geomarkets remained consistent through 2020, a slight median income increase for the City of Chicago Geomarket (\$71,000) and decrease in the South and Southwest Suburbs Geomarket (\$84,000) narrowed the income gap between the lowest and second-lowest Geomarkets to just \$13,000. @fig-chicago-rq1-ses also shows that from 1980 to 2020, the percentage of residents living in poverty declined in the City of Chicago Geomarket while it increased in all other Geomarkets, resulting in a narrowing of poverty gaps across the Chicago metropolitan area. Similar patterns are evident in educational attainment across Geomarkets. In 1980, the City of Chicago and South and Southwest Suburbs had the lowest percentage of residents with a bachelor's degree or higher (14%). However, by 2020, the City of Chicago experienced the largest percentage point increase in college degree attainment, reaching 34%.

>>>>

On January 9, 2019 a public research university purchased College Board Student Search Service order no. $448922$, thereby obtaining the contact information of out-of-state prospective students, who would subsequently receive brochures, postcards, emails, and targeted social media. This was a large student list purchase, yielding $122,426$ "names." At $\$0.45$ per name, the order cost $\$55,091.70$. The order specified three "search filters" to control which names were purchased: first, the 2020 high school graduating class; second, under the heading "College Board Exams," PSAT score range of 1070 - 1180; and third, under the heading "Geography", the order filtered on 67 "Geomarkets," identified by titles like "IL08 - Northwest Suburbs" and "IL09 - North Shore."[^1] The Geomarket filter carves states and metropolitan areas into smaller geographic units. Upon closer inspection, this student list purchase selected all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago," which is significantly poorer and has a higher share of Black residents. Similarly, in the DC/Maryland/Virginia metro area, the order selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington and Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets.

@@

**Correlation**. Broadly, correlation measures the extent to which two or more variables move together. Predictive algorithms are based on correlation and are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and/or to assign levels of risk to each case. @RN4975 provides the example of @RN4989, who developed a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 [p. 50] writes that, "correlation grounds big data's so-called revolutionary potential. As Wired editor Chris Anderson infamously declared\ldots, big data proved that 'correlation supersedes causation, and science can advance even without coherent models, unified theories, or really any mechanistic explanation at all.'"

@@

Negative consequences of homophily
Chun (2021) discusses the negative consequences arising from homophily.Second, when we "engineer homophily" – through algorithms or policies – we make it less likely for heterophily to emerge, eliminating possibilities – of friendships, of matches – that may have occurred in the absence of such engineering.

@@

**Geomarket Filter in Student Search Service**. <!-- By 1984, College Board Geomarkets were incorporated as a search filter in the College Board Student Search Service \[CITE\]. --> Geomarkets also function as a student list search filter for the College Board's Student Search Service. A student list contains the contact information of prospective students who meet the search filter criteria (e.g., test score, GPA) specified by the university. Student lists are the fundamental input for undergraduate recruiting campaigns because purchased names -- alongside prospects who reach out on their own -- constitute the set of prospects who receive subsequent recruiting interventions (e.g., mail, email) designed to push them toward the application and enrollment stages of the "enrollment funnel." @RN4895 reports that 86% of public colleges and 87% of private colleges purchase student lists.[^11] Historically, dominant list vendors are the College Board and ACT, which derived student list data from test-takers, but the test-optional movement, advances in technology, and surging private equity investments have contributed to new sources of student list data [@list_biz].

@@

**RQ1**. RQ1 primarily utilizes Census Data while RQ2 primarily utilizes data about student list purchases. Research question 1 is about socioeconomic and racial variation between Geomarkets in metropolitan areas and how such variation changes over time. To answer this question, we use census tract data from the U.S. Census Bureau. Utilizing a small geographic area is important for implementing a spatial merge between Census data and Geomarket shape files (described below). Census tracts are the smallest geographic area for which measures of both race/ethnicity and socioeconomic characteristics are available. Defined as statistical subdivisions of counties, census tracts boundaries are reassessed every decennial census to reflect relatively homogeneous demographic communities, typically ranging in population size from 2,500 to 8,000 residents [@Bureau2014; @Bureau1994]. Compared to zip codes, which are fixed spatial units designed primarily to demarcate route efficiency for postal services, census tracts are considered more reliable for population-based analyses.

@@

Across these data sources, we created measures of race and ethnicity, specifically the number and proportion of the census-tract population in the following categories: White, non-Hispanic; Black, non-Hispanic; Hispanic; Asian, non-Hispanic; Native Hawaiian and Pacific Islander, non-Hispanic; American Indian and Alaskan Native, non-Hispanic; and two or more races, non-Hispanic. Measures of Asian, Native Hawaiian and Pacific Islander, American Indian and Alaskan Native, and two or more races (all non-Hispanic) were not available for the 1980 Decennial Census. We also created socioeconomic measures that approximate the predictors of market segment identified by @RN4982. These variables are median household income, percent of households below the poverty line, and people age 25+ with a BA. All measures were created at the census tract-level and at the Geomarket-level by aggregating tract-level data to Geomarkets.

@@

[^17]: An alternative approach to dealing with spatial misalignment included creating shapefiles for Geomarkets from a 2012 R-bloggers post written by a data scientist working in the enrollment management space. Essentially, the data file was a Geomarket-zip code crosswalk that held one observation per zip code and a column that assigned each zip code to one Geomarket. Using these zip code-level data, we utilized the aggregate() function to create shapefiles for each Geomarket. Next, we utilized the st_intersection() function to assign each census tract to a Geomarket, based on the intersection of their associated spatial shapefiles. This spatial merge kept Geomarket boundaries consistent across all three Decennial Census and split census-tract shapefiles that intersected multiple Geomarkets into smaller shapefiles, each entirely contained within a single Geomarket. We replaced each variable in these shapefiles with its original value scaled by the share of the tract’s land area that lay within the Geomarket. For a particular census tract, imagine that 60% of its land area was contained in Geomarket A, 40% of its land area was contained in Geomarket B, and the census tract reports 100 people who identify as Asian, non-Hispanic. The partial spatial merge splits this census tract into two observations. The observation assigned to Geomarket A reports $.60*100 = 60$ Asian, non-Hispanic people and the observation assigned to Geomarket B reports $.40*100 = 40$ Asian, non-Hispanic people. Even under this most conservative approach to addressing spatial misalignment, overall patterns in the findings remained consistent. However, we opted to georeference the manually digitized maps directly sourced from the College Board, as the original source of the Geomarket-zip code crosswalk cited in the R-bloggers post could not be verified.

@@

The data collection sample for this project was all public universities in California, Illinois, Minnesota, and Texas. We chose these states because they were priorities for the two foundations that funded the data collection. Utilizing public records requests to obtain quantitative data is a painstaking process, made more difficult by the onset of the COVID-19 Pandemic. Initially, the majority of universities ignored or denied our request. We established a client relationship with a civil rights legal organization and leveraged this relationship to obtain pro bono representation from four law firms. We also collected data from Arizona State University and Northern Arizona University, because we were able to obtain pro-bono legal representation for these two universities. However, we were unable to obtain legal representation for Texas because firms interested in representing us had conflicts of interest in Texas.

@@

The variables of interest are Geomarket and prospect-level characteristics, which are derived from the pretest questionnaire administered to College Board test-takers that are included in the de-identified student lists we received (e.g., HS code, HS GPA range, intended major). College Board asks students separate questions about ethnicity (Cuban, Mexican, Puerto Rican, other Hispanic, non-Hispanic, ethnicity non-response) and race (American Indian or Alaska Native, Asian, Black, Native Hawaiian or other Pacific Islander, White, race non-response), allowing students to check as many boxes as they want [@RN0010]. From these responses, a College Board "derived aggregate race/ethnicity" was re-created based on the U.S. Department of Education reporting guidelines that includes the following categories: no response; American Indian/Alaska Native; Asian; Black; Hispanic/Latino; Native Hawaiian or Other Pacific Islander; White; other; two or more races, non-Hispanic [@RN0010].[^20] Information about student socioeconomic status is limited. However, we create a measure of whether the prospect is a first-generation college student based on their parents' level of education that was self-reported on the College Board questionnaire.

[^20]: Any student who selects a Hispanic ethnicity category is defined as Hispanic/Latino, regardless of the race categories they select. Additionally, non-Hispanic students who check "American Indian or Alaska Native" and another race group are defined as "two or more races, non-Hispanic."

@@

**Research design**. We utilize a multiple, quantitative case study design. Metropolitan areas are cases. Case studies deepen understanding of phenomena by comparing and contrasting across different settings [@RN4011]. Our research questions explore the socioeconomic and racial variation between Geomarkets, how such variation changes over time, and how Geomarket landscapes shape the composition of included versus excluded prospects. Our goal is not to make generalizations about Geomarket variation nationally, but to explore how different geographical models of metropolitan areas, shaped by their unique demographic characteristics, influence the composition of prospect pools. Such aims align with the purpose of case studies, which drive the exploration of research questions focused on the "how" and "why" of phenomena in rich and nuanced contexts. We, therefore, select a small number of metropolitan areas in order to provide sufficient analytic depth while situating each case within its specific historic and contemporary sociodemographic characteristics. The [Online Appendix](https://ozanj.github.io/cb_geo/#/results) presents results for a larger number of metropolitan areas.

The central thesis of Zemsky & Oedel's (1983) Market Segment Model suggests that student demand for various colleges is a function of social class and geography, and that colleges strategically target recruiting in high socioeconomic areas. However, because Geomarkets vary significantly in their racial and class characteristics, and in how colleges target these Geomarkets, a multiple case study enables us to compare how these dynamics unfold across different metropolitan areas. By treating each metropolitan area as a separate case, we explore both within-case processes (e.g., how homophily structures recruitment in a single metropolitan area) and cross-case patterns (e.g., how the same recruitment strategy results in different racial/class outcomes across metropolitan areas).

Case selection was informed by purposeful sampling to identify "information-rich" cases that can provide knowledge about variation across Geomarkets and over time [@RN4011; @RN4549]. We considered metropolitan areas as information rich cases across four broad dimensions: 1) urban model structures, 2) geographical variation, 3) variation across racial/ethnic and socioeconomic characteristics, and 4) availability of student list orders and de-identified list data. Based on @RN4982, we identified classic urban model structures that reflected "major metropolitan area\[s\] \ldots composed of several markets, usually corresponding to the inner city, a first ring of suburbs, and an outer ring of suburbs" (p. 11-12), such as Boston, Chicago, Detroit, Baltimore, Denver, Houston, Miami, and Philadelphia. To compare and contrast with classic urban model structures, we also identified alternative urban models such as bi-modal or metroplex models that contain two centralized cores surrounded by the more classic suburban ring (e.g., Dallas-Fort Worth and San Francisco) and multi-nuclei models where urban growth develops around multiple, sometimes uncentralized and independent, metropolitan areas (e.g., Los Angeles and New York City). The third consideration addresses geographical variations in demographic populations by intentionally selecting metropolitan cases across different U.S. regions (West, Midwest, Northeast, South) to ensure resident and prospect pools mirror broader populations. Related to all previous selection dimensions, we are also interested in variation of racial and socioeconomic populations between Geomarkets within a metropolitan area (RQ1) and how such variation shapes which prospective students are targeted by student list purchases (RQ2). We therefore considered whether or not metropolitan areas provided sufficient and varied racial and socioeconomic diversity for our research purposes.

Lastly, choice of metropolitan cases was informed by student list data availability. Although Census data are available for all metropolitan areas nationally (RQ1), we do not have good candidate student list data for all metropolitan areas (RQ2). In order to show which prospective students would have been included or excluded from the student list purchase had the purchase filtered on particular Geomarkets, the student list purchases we analyze must include all Geomarkets in a metropolitan area. This lead us to select on orders that filtered for an entire state (e.g., State filter is set to Pennsylvania), an entire metropolitan area without a Geomarket filter (e.g., CBSA filter is set to the Philadelphia metropolitan area), or an entire metropolitan area with a Geomarket filter inclusive of all Geomarkets (e.g., CBSA filter is set to the Philadelphia metropolitan area and Geomarket filter is set to all five Geomarkets in the Philadelphia area). Although we would like to analyze the D.C./Maryland/Virginia (DMV) area, the student list purchases we collected exclude "MD 5 -- Prince George's Metropolitan" and "MD 7 -- Baltimore (Urban)," which are substantially poorer and more Black than other Geomarkets in the DMV. Similarly, student list purchases targeting Cleveland exclude "OH 4 -- City of Cleveland (East)."

@@

We answer RQ2 by analyzing student list purchases that encompassed all Geomarkets in a selected metropolitan area. Descriptive tables and graphs describe the racial and socioeconomic characteristics of prospects that would have been included/excluded had the purchase filtered on particular Geomarkets. We also create interactive maps of prospect pools by Geomarket contributions to provide a more granular visualization of analyses.

@@

@fig-chicago-maps presents the seven Geomarkets in the Chicago metropolitan area by total population across the three census periods. The maps in the figure show how Chicago Geomarkets reflect the classic urban geographical structure of an inner-city core surrounded by a ring of suburbs as described by @RN4982 [p. 12]. Based on the color gradient scale representing total population density, @fig-chicago-maps indicates northern suburb Geomarkets in the metropolitan area -- Chain of Lakes, Northwest Suburbs, Northshore, Evanston and Skokie -- experienced relatively little population growth from 1980 through 2020. In contrast, the most significant population increases occurred in the Western Suburbs (1.0 Million to 1.4 Million) and the South and Southwest Suburbs (less than 1 Million to 1.2 Million) Geomarkets. The City of Chicago Geomarket, forming the metropolitan area's central core, was the only locality to experience a population decline, from 3.5 Million in 1980 to 3.2 Million in 2020, despite remaining the most populous Geomarket.

These modest population shifts belie substantial change in the racial/ethnic composition within and across Geomarkets over time. @fig-chicago-rq1-race illustrates the racial and ethnic composition of each Geomarket across the three census periods. In 1980, nearly all Geomarkets were predominantly White, with proportions ranging from 84% in the South and Southwest Suburbs to approximately 96% in North Shore. In contrast, the City of Chicago had a substantially more diverse population, with White residents comprising only 52% of its total population. One data caveat is that the following race/ethnicity categories were not available in 1980: Asian, non-Hispanic; Two+ races, non-Hispanic; NHPI, non-Hispanic; AIAN non-Hispanic. By 2020, several Geomarkets that were previously predominantly White shifted to a more heterogeneous racial/ethnic distribution, including Chain of Lakes, Northwest Suburbs, Evanston and Skokie, Western Suburbs, and South and Southwest Suburbs. For many Geomarkets, this balance in racial/ethnic composition was the result of large increases in the Latinx population from 1980 to 2020, which ranged from a 7 percentage point increase for the Evanston and Skokie Geomarket to a 20 percentage point increase for Chain of Lakes. Some Geomarkets also experienced a substantial increase in Black populations (South and Southwest Suburbs) and in Asian populations (Northwest Suburbs and Evanston and Skokie). @fig-chicago-rq1-race also indicates that, although the North Shore Geomarket underwent some changes in racial and ethnic composition, its overall demographic patterns remained relatively the same across the three census periods. City of Chicago experienced relatively modest declines in its Black population (35% in 1980 to 26% in 2020), alongside increases in its Hispanic (13% in 1980 to 30% in 2020) and Asian populations (4% in 2000 to 7% in 2020), further substantiating the Geomarket as the most racially/ethnic diverse in the metropolitan area. <!-- For instance, North Shore experienced some increases in Hispanic (2% in 1980 to 8% in 2020), Asian Pacific Islander (7% in 2000 to 10% in 2020), and Black (2% in 1980 to 3% in 2020) populations yet remained a predominantly White Geomarket (74% in 2020). City of Chicago also experienced relatively modest declines in its Black population (37% in 1980 to 33% in 2020), alongside increases in its Hispanic (15% in 1980 to 26% in 2020) and Asian Pacific Islander populations (4% in 2000 to 6% in 2020), further substantiating the Geomarket as the most racially/ethnic diverse in the metropolitan area. -->

The interactive maps in the [Online Appendix](https://ozanj.github.io/cb_geo/#/research-question-1-15) can show these shifts at the census tract-level. For example, in 1980 Black people were concentrated in southern tracts of the City of Chicago Geomarket, as described by scholarship on gentrification in Chicago [@ewing2024_nofuture; @snidal2022_exodus]. However, in 2000 and 2020 Black populations shift towards northeast tracts of the South and Southwest Suburbs Geomarket

We explore socioeconomic demographics -- median income, percent living below poverty, and educational attainment -- within and across Geomarkets over time in @fig-chicago-rq1-ses. The primary takeaways are the City of Chicago -- and to a somewhat lesser degree the South and Southwest Suburbs -- stands out as having lower median income, higher poverty rates, and lower BA attainment than other Geomarkets. By contrast, the North Shore stands out as being particularly affluent and highly educated. Overall, median income both within and across Geomarkets in the Chicago metropolitan area remained relatively stable across the three census periods. For instance, in 1980, median household income across the Chicago metropolitan area ranged from a high of \$165,000 in the North Shore Geomarket to a low of \$68,000 in the City of Chicago Geomarket. However, median incomes in all other Geomarkets were closer to the upper thresholds of affluence, such as \$125,000 in Evanston and Skokie, \$117,000 in the Northwest Suburbs, \$108,000 in the Western Suburbs, \$104,000 in Chain of Lakes, and \$99,000 in the South and Southwest Suburbs. While overall income patterns across Geomarkets remained consistent through 2020, a slight median income increase for the City of Chicago Geomarket (\$71,000) and decrease in the South and Southwest Suburbs Geomarket (\$84,000) narrowed the income gap between the lowest and second-lowest Geomarkets to just \$13,000. @fig-chicago-rq1-ses also shows that from 1980 to 2020, the percentage of residents living in poverty declined in the City of Chicago Geomarket while it increased in all other Geomarkets, resulting in a narrowing of poverty gaps across the Chicago metropolitan area. Similar patterns are evident in educational attainment across Geomarkets. In 1980, the City of Chicago and South and Southwest Suburbs had the lowest percentage of residents with a bachelor's degree or higher (14%). However, by 2020, the City of Chicago experienced the largest percentage point increase in college degree attainment, reaching nearly 40%.
